Higgs Boson Machine Learning Challenge
========================================================

The goal of this challenge is to identify which simulated ATLAS experiment proton-proton collision
observations contain evidence of boson pairs ("signal"), and which do not ("background"). 

The challenge is [hosted by Kaggle](http://www.kaggle.com/c/higgs-boson) and provided by [The Atlas Experiment](http://atlas.ch/).  Technical details are found in the [Laboratoire de l’Accélerateur Linéaire](http://higgsml.lal.in2p3.fr/) technical documentation [PDF](http://higgsml.lal.in2p3.fr/files/2014/04/documentation_v1.5.pdf)

# The Data

Provided were a training set of 250,000 observations and a test set of 500,000 observations.
Unusable or unknown quantities are reported as -999.0.

The variables include 17 "primitive" quantities and 13 "derived" quantities. Also included
is the EventId identifying the observation. The training set also includes the goal label (b or s)
and a "Weight" metric not to be used in classification. For this reason I've stripped it out 
of the training set and saved a new CSV as processed_train.csv

## The Variables
```{r load_data, cache=TRUE}
    train <- read.csv("data/processed/processed_train.csv")
    names(train)
```
```{r load_libraries}
    require(ggplot2)
```
```{r plot_function}
    create_plot <- function(data, v) {
        p <- ggplot(data, aes(x=EventId, y=data[,v], color=Label), environment=environment())
        l <- p + labs(y=v)
        print(l + geom_point())
    }
```


### Primitives

**PRI_tau_pt**

The transverse momentum $\sqrt{p^2_x+p^2_y}$, or the momentum perpendicular to the beam line, of the hadronic tau measured in GeV

```{r pri_tau_pt_fig, fig.width=8, cache=TRUE}
    create_plot(train, "PRI_tau_pt")
```
```{r pri_tau_pt}
    ptp <- train$PRI_tau_pt
    summary(ptp)
```
Most observations are < 175 GeV, let's replot on that subset
```{r subset_ptp_175, fig.width=8, cache=TRUE}
    ptp.175 <- subset(train, PRI_tau_pt < 175)
    create_plot(ptp.175, "PRI_tau_pt")
```
It looks like the slower the transverse momentum, the more likely the observation is background noise. Let's replot for < 50
```{r subset_ptp_50, fig.width=8, cache=TRUE}
    ptp.50 <- subset(train, PRI_tau_pt < 50)
    create_plot(ptp.50, "PRI_tau_pt")
```
**PRI_tau_eta**

The pseudorapidity η of the hadronic tau.







